---
title: "Homework 3"
subtitle: "Genoa Sullaway"
output: 
  word_document:
  fig_crop: no
---
----- builds off the HWK 3 script, started this after I turned in HW 3. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(here)
library(tidyverse)
library(lubridate)
library(sf)
library(maps)
library(viridis)
library(raster)
library(corrplot)
library(knitr)
library(mapproj)
library(sp)
library(mgcv)
library(GGally)
library(viridis) 
library(mapproj)
library(mgcViz) #helpful for model plotting and checking::
library(ggpubr)
#https://cran.r-project.org/web/packages/mgcViz/vignettes/mgcviz.html
#library(kableExtra)
theme_set(theme_classic())
#install.packages('terra', repos='https://rspatial.r-universe.dev')
```

```{r LOAD DATA}
#load data that was created by : final_DF_EDA_StatsProj.R
df <- read_csv(here("data","fulldata_stats_proj.csv"))  %>%
  filter(YEAR %in% c(2014, 2015, 2016, 2018)) %>%
   filter(!is.na(TEMPERATURE)) %>%
   mutate(MONTH = str_pad(MONTH, 2, pad = "0"),
         DAY = str_pad(DAY, 2, pad = "0")) %>%
   tidyr::unite(col = "date", c("YEAR", "MONTH", "DAY"), remove = FALSE, sep = "/")  %>%
  mutate(date = as.Date(date, "%Y/%m/%d"),
         julian = yday(date))
  

df_summary<- df %>% 
  mutate(TAXON_NAME = case_when(TAXON_NAME %in% c("Calanus marshallae", "Calanus pacificus", "Calanus glacialis") ~ "Calanus spp",
                                TRUE ~ TAXON_NAME)) %>%
  group_by(YEAR, MONTH, julian,LAT, LON, TEMPERATURE, SALINITY, DOMAIN, MAX_GEAR_DEPTH, BOTTOM_DEPTH,TAXON_NAME, summer.cold.pool.extent) %>%
  dplyr::summarise(sum = sum(EST_NUM_PERM3))

#convert from lat long to x y albers. 
x <- mapproject(df_summary$LON, df_summary$LAT, "albers", param=c(55.9, 60.8))
df_summary$x <- x$x #LON
df_summary$y <- x$y #LAT 
 
# climate_data <- read_csv("data/climate_data.csv") %>%
#   filter(year %in% c(2014, 2015,2016,2018)) %>%
#   dplyr::select(year, summer.cold.pool.extent)  
# 
# summer_CP <- data.frame(unique(climate_data[c("year", "summer.cold.pool.extent")]))
```

```{r create base maps shape file,include=FALSE}
#create base map
ak <- sf::st_as_sf(maps::map('world','USA:Alaska',
                             # ylim = c(55,60),
                             #   xlim=c(-140,-130), 
                             plot=FALSE, fill=TRUE))

#create bounds to trim AK map 
bounds_ak <- extent(-176,-156,54,65)
#bounds_ak <- extent(-164.5,-160, 55,60) 
extent_ak <- st_set_crs(st_as_sf(as(bounds_ak, "SpatialPolygons")), 4326)
ak <- st_intersection(ak, extent_ak) #trim map by intersections 
 plot(ak)
# EPSG = 5070
ak_albers<- st_transform(ak,5070) #change to albers projection
plot(ak_albers)
 m
#load shape file with ortiz shapes 
ortiz_shp <- st_read("data/ortiz_regions/BSIERP_regions_2012.shp")
ortiz_shp <- subset(ortiz_shp, !DOMAIN == 15)
#Tranform into WGS84 coordinate system
ortiz_shp <- st_transform(ortiz_shp, "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs")
 
#  sf::st_as_sf(maps::map('world','USA:Alaska',
#                             # ylim = c(40,75),
#                               xlim=c(-175,-115), 
#                              plot=TRUE, fill=FALSE))
```

## Title: 
Evaluating spatial variability in *Calanus marshallae* distribution during the 2014-2016 Marine Heatwave in the Eastern Bering Sea

## Introduction/Background:
Marine heatwaves (MHW) have become more prevalent as anthropogenic climate change progresses (Frolicher et  al. 2018). In  the Northeast Pacific Ocean, the most recent MHW (2014-2016) resulted in mass mortalities  and marine mammal strandings, species distribution shifts, fishery closures, and social-ecological network shifts (Cheng and Frolicher 2020, Fisher et al. 2021). 

Primary and secondary production are tightly linked to oceanographic conditions and are thus heavily influenced by MHW's. In the Eastern Bering Sea (EBS), copepods such as *C. marshallae* underpin the structure of the entire ecosystem. As high nutrient prey items, *C. marshallae* support marine mammals, numerous subsistence fisheries and a billion-dollar Pollock fishery. However, copepod abundance and community composition are heavily regulated by environmental factors (Siddon et al 2013). During cold years, large nutritious copepods are abundant into the Fall and are key prey items for pollock (Siddon et al 2013), seabirds (Springer et al 1986) and marine mammals (Baumgartner et al. 2013). *C. marshallae* abundance is tightly linked with pollock recruitment through the Oscillating Control Hypothesis (Kimmel et al. 2018). In warm stanzas, the zooplankton community is  dominated by smaller species with lower lipid content (Kimmel et al 2018) often resulting in poor pollock recruitment. This disruption in lipid-rich prey during warm years has negative effects that cascade throughout the EBS food webs and social-ecological networks. 

Large scale climate events can lead to increases in spatial synchrony across increasingly large distances (Hanson et al. 2020). MHW's may increase synchrony in *C. marshallae* abundance across the EBS, and this can have cascading negative effects through the ecosystem. Moreover, changes to spatial variability in *C. marshallae* within warm years can have implications for future predator-prey overlap (Siddon et al 2013). This research aims to understand the spatial dynamics of *C. marshallae* within warm years, and (if time allows) the impact of successive warms years on abundance. 

## Goals and Objectives

The research question I would like to address is: Is there spatial variation in *C. marshallae* abundance during the 2014-2016 Marine Heatwave in the Eastern Bering Sea?  

Warm stanzas and marine heatwaves are predicted to become more common in the EBS. I want to understand the spatial variability for *C. marshallae* among warm years, to ultimately inform the spatial availability of *C. marshallae* prey for higher trophic levels. Does the EBS become a "food desert" for higher trophic levels or are there certain spatial refuges during warm years? 

Generally, copepod abundance and phenology are tightly linked to oceanographic conditions. I expect variation in temperature to be a strong predictor in *C. marshallae* CPUE. Specifically, I expect a negative linear or second order polynomial relationship between temperature and *C. marshallae* abundance. I have not found evidence of a non-linear relationship like this in the literature, but I could imagine a threshold effect where temperature gets too high and *C. marshallae* abundance is consistently low past a certain high temperature threshold. Additionally, I expect an effect of year as the MHW progresses, ie. I expect 2014 to be different than 2016 because of cumulative impacts of warm years. I expect *C. marshallae* to be more abundant in the northern inner shelves compared to the southern and outer shelves, but mediated by temperature (Kimmel et al 2018).

Proportional wind direction has a significant impact on the  Bering Sea ecosystem (Danielson et al 2012), I expect a positive linear relationship between the proportion of NE wind and *C. marshallae* CPUE, because NW wind encourages colder conditions and southern ice flow. Additionally, I expect a negative linear relationship between proportion of SE wind and  *C. marshallae* CPUE, however I expect SW winds to be more prevalent during the MHW years that I am examining. 

## Data Description & Collection Methods

Data were collected by NOAA AFSC ECOFOCI during at-sea oceanographic cruises in the Eastern Bering Sea (Figure 1). Zooplankton data were provided by Dr. Dave Kimmel. The response variable is *Calanus marshallae* catch per unit effort (CPUE) (individuals m^3) for each station (each point in Figure 1) sampled in each year of the recent MHW (2014-2016). 

Zooplankton data are collected annually on NOAA oceanographic cruises from April-October, however the spatial and temporal schedule is unbalanced from year to year (Figure 1). Copepods are collected using paired bongo nets (20cm frame, 150um mesh and a 60cm frame, 333 um mesh) on oblique tows. The tow depth varies by station and is not necessarily related to bottom depth. If bottom depth is less than 100m, max tow depth is 5-10m above the bottom. However, if max bottom depth is deeper than 100m the max gear depth can vary depending on the sea state, most tows are not deeper than 300m. The samples are identified to lowest taxonomic level and sorted into different stages (C-1 to C-6), however for this analysis I have summed *C. marshallae* CPUE across life stages. 

Some covariate data is collected onboard the ship, for example maximum bottom depth and gear depth are recorded on the zooplankton gear deployment. Salinity and temperature covariates were collected from CTD casts done in parallel to zooplankton net casts. 

Covariates such as proportional daily wind direction and cold pool extent were downloaded from publicly available climate dataset provided by Dr. Mike Litzlow (NOAA AFSC, https://github.com/mikelitzow/bold-new-pollock/tree/master/data). The proportion of daily wind from the SE/NW from May to September is used as a simple index for changes in Bering Sea advection (Danielson et al 2012). Summer cold pool extent is calculated on the NMFS bottom trawl survey. Finally, we use Domain as a covariate to represent the Ortiz Regions. These regions have been established based on known differences across oceanographic breaks in the Bering Sea.


Exploratory data analysis demonstrate that there is unbalanced sampling of stations across years (Table 1, Figure 1). Exploratory plots indicate spatial variability in *C. marshallae* abundance and temperature throughout the EBS  (Figure 1 and 2). A coplot of select covariates indicate possible relationships between depth, salinity and temperature. We see that there is more variation in temperature and salinity at shallower depths across years (Figure 3). It also appears there is generally less variation in temperature at depth, which is expected based on the differences in water mixing. Interestingly, temperatures at depth is warmer and less variable as the MHW progresses (Figure 3).

\n

```{r data table }
library(knitr)

temp <- unique(df[c("YEAR", "STATION_NAME")])

station_year <-temp %>% 
   mutate(id=1) %>%
   group_by(YEAR) %>%
   dplyr::summarise(value = sum(id)) %>%
   dplyr::rename(variable = "YEAR") %>%
   mutate(temp = "Stations surveyed per year, ") %>%
   dplyr::select(temp, variable, value) %>%
   tidyr::unite("variable", 1:2, sep = "")
  

depth <- df %>%
  dplyr::summarise(Min_Gear_Depth = min(MAX_GEAR_DEPTH), 
            Max_Gear_Depth = max(MAX_GEAR_DEPTH)) %>%
   gather(1:2, key = "variable", value = "value") %>%
   mutate(value = paste(value, collapse=" - ")) %>%
   mutate(variable = "Survey Depth m (Min, Max)") %>%
   slice(1) 
 
gear <-df %>%
  summarise(value = unique(GEAR_NAME)) %>%
  mutate(variable = "Gear Type") %>%
  mutate(value = paste(value, collapse=", ")) %>%
  slice(1) 

salinity <- df %>%
  dplyr::summarise(Min_Salinity = round(min( SALINITY,  na.rm = T)),
            Max_Salinity = round(max(SALINITY, na.rm = T))) %>%
   gather(1:2, key = "variable", value = "value") %>%
   mutate(value = paste(value, collapse=" - ")) %>%
   mutate(variable = "Salinity ppm (Min, Max)") %>%
   slice(1)

temperature <- df %>%
  dplyr::summarise(Min_Temp = round(min( TEMPERATURE,  na.rm = T)),
            Max_Temp = round(max(TEMPERATURE, na.rm = T))) %>%
   gather(1:2, key = "variable", value = "value") %>%
   mutate(value = paste(value, collapse=" - ")) %>%
   mutate(variable = "Temperature C (Min, Max)") %>%
   slice(1)

# wind.nw <- df %>%
#   dplyr::summarise(Min_NW_Wind = round(min(NW.wind.May.Sep,  na.rm = T), digits = 2),
#             Max_NW_Wind =  round(max(NW.wind.May.Sep, na.rm = T), digits = 2)) %>%
#    gather(1:2, key = "variable", value = "value") %>%
#    mutate(value = paste(value, collapse=" - ")) %>%
#    mutate(variable = "Propotion Daily NW Wind (Min, Max)") %>%
#    slice(1)

# wind.se <- df %>%
#   dplyr::summarise(Min_SE_Wind = round(min( SE.wind.May.Sep,  na.rm = T), digits = 2),
#             Max_SE_Wind = round(max(SE.wind.May.Sep, na.rm = T), digits = 2)) %>%
#    gather(1:2, key = "variable", value = "value") %>%
#    mutate(value = paste(value, collapse=" - ")) %>%
#    mutate(variable = "Proportion Daily SE Wind (Min, Max)") %>%
#    slice(1)

# coldpool <- df %>%
#   dplyr::summarise(Min_Cold_Pool = round(min(summer.cold.pool.extent,  na.rm = T)),
#             Max_Cold_Pool = round(max(summer.cold.pool.extent, na.rm = T))) %>%
#    gather(1:2, key = "variable", value = "value") %>%
#    mutate(value = paste(value, collapse=" - ")) %>%
#    mutate(variable = "Summer Cold Pool Extent km (Min, Max)") %>%
#    slice(1)
 
 table <-rbind(station_year, depth, gear, salinity,temperature#,wind.se,wind.nw,coldpool
               ) %>%
  dplyr::rename(Variable = "variable", Value = "value")

# Generally, the table and figures show the key features of the data. One improvement would be to list means and ranges of sampling dates and covariates by year (3 columns for 2014,15,16)
 
table_summ <- df_model %>%
  filter(TAXON_NAME %in% c("Calanus spp", "Pseudocalanus spp")) %>%
  group_by(TAXON_NAME, YEAR,MONTH ) %>%
  summarise( mean = mean(sum))

table_summ2 <- df_model %>%
  filter(TAXON_NAME %in% c("Calanus spp", "Pseudocalanus spp.")) %>%
  group_by(TAXON_NAME, YEAR ) %>%
  summarise( mean = mean(sum),
             mean_temp = mean(TEMPERATURE),
             Min_Temp = round(min( TEMPERATURE,  na.rm = T)),
            Max_Temp = round(max(TEMPERATURE, na.rm = T)))
```

\n
Is there multicolinearity??
- I think this plot says yes... esp with salinity and temperature, also salinity and depth

```{r covariate}
df <- df_summary %>%
  filter(!SALINITY < 29)
ggpairs(df[,c("sum", "TEMPERATURE", "MAX_GEAR_DEPTH")], lower=list(continuous='smooth')) #est num, salinity, temp, depth
 
#wind: ,31:32

#coldpool -- 33 

ggplot(df_model,aes(x=log(sum), y = TEMPERATURE)) +
  geom_point( ) +
  geom_smooth() + 
  facet_wrap(~YEAR)

ggplot(df_model,aes(x=log(sum), y = MAX_GEAR_DEPTH)) +
  geom_point( ) +
  geom_smooth() +
  facet_wrap(~YEAR)

ggplot(df_model,aes(x=log(sum), y = log(MAX_GEAR_DEPTH))) +
  geom_point( ) +
  geom_smooth()  

ggplot(df_model,aes(x=log(sum), y = SALINITY)) +
  geom_point( ) +
  geom_smooth() +
 facet_wrap(~YEAR)

ggplot(df,aes(x=log(sum), y = SALINITY)) +
  geom_point( ) +
  geom_smooth()  



```

\n

```{r }
kable(table,caption = "Table 1. Table of data structure")
```


```{r, eval = FALSE}

table<-df %>% 
  slice(1:5) %>%
  mutate()
  dplyr::select(CRUISE, BOTTOM_DEPTH,YEAR, MONTH,DAY,GEAR_NAME,MAX_GEAR_DEPTH,STATION_NAME, EST_NUM_PERM3, LAT, LON,SALINITY, TEMPERATURE, SE.wind.May.Sep, NW.wind.May.Sep, summer.cold.pool.extent)

kable(table) %>%
  kable_styling(font_size = 7)
```

\n
\n
I will need to test for multicolinearity across covariates, it is likely I do not need all the covariates I propose below. For example, depth, temperature and Ortiz region may be confounded since Ortiz regions were delineated based on oceanographic bounds and shelf breaks that differ by depth and temperature.  

\n

```{r PLOT ALL DATA, fig.cap ="Figure 1. *C. marshallae* log CPUE (individuals m^3) in the Eastern Bering Sea, where a point indicates a station where a survey was conducted,\n each facet represents a year, color indicates the log CPUE of *C. marshallae* at each station in each  year", fig.width=18, fig.height=8}
 #plot abundances for each station/time point. Color by Domain. 

cal_point_map<-ggplot(ak) +
  geom_sf() +
  geom_point(data = df_summary %>% filter(TAXON_NAME == "Calanus spp"), aes(x=LON, y=LAT, color = log(sum))) + #, size = EST_NUM_PERM3))+
  theme_classic() +
  scale_color_viridis() +
  labs(color= "C. marshallae Log(CPUE)")  +
  #scale_color_gradient(low = "#de9b71", high = "#5d74a5", name = "Log C. marshallae CPUE") +
  #guides(color=guide_legend(title="Log(CPUE)")) + 
  theme(strip.text = element_text(size =12),
       # legend.key.size = unit(2, 'cm'),
        legend.position = "bottom", 
        #legend.text=element_text(size=15),
       # legend.title = element_text(size=15),
        plot.caption = element_text(hjust = 0, size =10),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  facet_wrap(~YEAR) #+
 # labs(caption = "Figure 1. A point indicates a station where a survey was conducted,\n each facet represents a year, color indicates the log CPUE for calanus marshallae for each station in each  year")

cal_point_map

ps_point_map<-ggplot(ak) +
  geom_sf() +
  geom_point(data = df_summary %>% filter(TAXON_NAME == "Pseudocalanus spp."), aes(x=LON, y=LAT, color = log(sum))) + #, size = EST_NUM_PERM3))+
  theme_classic() +
  scale_color_viridis() +
  labs(color= "Pseudocalanus spp Log(CPUE)")  +
  #scale_color_gradient(low = "#de9b71", high = "#5d74a5", name = "Log C. marshallae CPUE") +
  #guides(color=guide_legend(title="Log(CPUE)")) + 
   theme(strip.text = element_text(size =12),
       # legend.key.size = unit(2, 'cm'),
        legend.position = "bottom", 
        #legend.text=element_text(size=15),
       # legend.title = element_text(size=15),
        plot.caption = element_text(hjust = 0, size =10),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  facet_wrap(~YEAR) #+
ps_point_map
```

 
\n

```{r PLOT TEMPERATURE, fig.cap ="Figure 2. Eastern Bering Sea temperature data (C) where a point indicates a station where a survey was conducted,\n each facet represents a year, color indicates the mean temperature at each station in each  year", out.width='\\textwidth',fig.align='center', fig.width=18, fig.height=8}
temperature<- ggplot(ak) +
  geom_sf() +
  geom_point(data = df, aes(x=LON, y=LAT, color = TEMPERATURE)) + #, size = EST_NUM_PERM3))+
  theme_classic() +
  scale_color_gradient(low = "blue", high = "red", name = "Mean Temperature (C) 100m") +
  labs(color= "Temperature")  +
  theme(  strip.text = element_text(size =12),
        legend.key.size = unit(2, 'cm'),
        legend.position = "bottom",    
        plot.caption = element_text(hjust = 0, size =10),
        legend.title = element_text(size=15),
        legend.text=element_text(size=15),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  facet_wrap(~YEAR) #+
#  labs(caption = "Figure 2. A point indicates a station where a survey was conducted,\n each facet represents a year, color indicates the mean temperature at each station in each  year")
temperature
```

\n

```{r PLOT julian day, out.width='\\textwidth',fig.align='center', fig.width=18, fig.height=8}
julian_day<- ggplot(ak) +
  geom_sf() +
  geom_point(data = df, aes(x=LON, y=LAT, color = julian)) + #, size = EST_NUM_PERM3))+
  theme_classic() +
  scale_color_gradient(low = "blue", high = "red", name = "Julian day") +
  labs(color= "Julian day")  +
  theme(  strip.text = element_text(size =12),
        legend.key.size = unit(2, 'cm'),
        legend.position = "bottom",    
        plot.caption = element_text(hjust = 0, size =10),
        legend.title = element_text(size=15),
        legend.text=element_text(size=15),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  facet_wrap(~YEAR) #+
#  labs(caption = "Figure 2. A point indicates a station where a survey was conducted,\n each facet represents a year, color indicates the mean temperature at each station in each  year")
julian_day
```

```{r julian day and temperature scatter plot}

ggplot(data = df) +
  geom_point(aes(x=julian, y = TEMPERATURE)) + 
  facet_wrap(~YEAR)

```

```{r plot coplot, fig.cap="Figure 3. Relationship between some covariates used in the analysis, \n temperature, salinity, maximum gear depth and year from\n stations where zooplankton were surveyed" }
#remove one salinity outlier 
# df_no_sal <- df %>%
#   filter(!SALINITY<23 )

coplot(TEMPERATURE~julian|LAT * as.factor(YEAR), data = df_calanus, number=c(4,3))
mysubtitle = "Figure 3. Coplot demonstrating the\nrelationship between temperature (C), \nsalnity (ppm), maximum gear depth (m) and year from\n stations where zooplankton were surveyed"
mtext(side=3, line= -28, at=-0.2, adj=0, cex=0.7, mysubtitle)

```

\n

```{r data histogram, fig.cap="Figure  4a. Log-transformed CPUE (individuals m^3) for the complete  *C. marshallae* dataset"}
df_summary %>%
  filter(TAXON_NAME == "Calanus spp") %>%
  ggplot() +
  geom_histogram(aes(sum), bins = 50)+
  xlab("Log CPUE") +
  ylab("Count") 

  #labs(caption = "Figure  4. Log-transformed CPUE for the complete  *Calanus marshallae* dataset")
# 
# test<- df_summary %>% 
#   filter(sum == 0) %>%
#   summarise(count(nrow()))
# 
# test<- df_summary %>% 
#   filter(sum > 0) %>%
#   filter(sum<1) %>%
#   summarise(count(nrow()))
# 
# test <- df_summary %>%
#   filter(!EST_NUM_PERM3 ==0) %>%
#   ggplot() +
#   geom_histogram(aes(EST_NUM_PERM3))+
#   xlab("Log CPUE") +
#   ylab("Count") 
# 
# df_nozero %>%
#   ggplot() +
#   geom_histogram(aes(log(sum))) + 
#   xlab("Log CPUE") +
#   ylab("Count") 
  #labs(caption = "Figure  4. Log-transformed CPUE for the complete  *Calanus marshallae* dataset")
```

\n

```{r by year data histogram, fig.cap="Figure  4b. Log-transformed *C. marshallae* CPUE (individuals m^3) by year"}
df %>%
  ggplot() +
  geom_histogram(aes(log(EST_NUM_PERM3) ))+
  facet_wrap(~YEAR)+ 
  ylab("Count") + 
  xlab("Log CPUE")
  #labs(caption = "Figure  4. Log-transformed CPUE for the complete  *Calanus marshallae* dataset")
```

## Methods

These data may have spatial autocorrelation and they may have a slight right skew (though this relationship changes when data are plotted by year). However, with a log transformation (Figure 4) data appear normal, so I propose to use a log normal distribution with a log link for the model. The random variable I would like to estimate is *C. marshallae* CPUE where $\gamma_i$ = Calanus m-3
  
$$log(\gamma_{s,y}) \sim Tweedie(\mu_{s,y},\sigma^2) $$

# explore gam, constrain smooth k 3-4. 
# if I want to include wind or CP add it one at a time, if I use CP dont need to include year
# 3-4 hypothesis based models

$log(\mu_{s,y}) = \alpha + a_t + b_sDomain_s + B_1Depth_{s,y} + B_2Temp_{s,y} + B_3SAL_{s,y} + B_4Wind_{nw,y} + B_5Wind_{se,y}+  B_6CP_y + \varepsilon_s$

$$\varepsilon_s = MVN(0,\sigma)$$

Where $b_{0s}$ is the random effect of year (note I am interested in looking at temporal autocorrelation, I think the order of the year matters as the heatwave progresses), $Depth_{s,y}$  is the survey depth, $Temp_{s,y}$ is the mean temperature (by survey depth) at each station for each survey, $SAL_{s,y}$ is the mean salinity (by survey depth) at each station for each survey, $Wind_{nw,y}$ is the proportional daily wind from the NW for each year, $Wind_{se,y}$ is the proportional daily wind from the SE for each year, $CP_y$ is the cold pool extent for that year, $Domain_s$ is a random effect of domain, this is based on the Ortiz regions delineated by oceanographic bounds. The error term, $\varepsilon_i$, is assumed to be multivariate normally distributed and $\sigma$ will be composed of a variance-covariance matrix that accounts for spatial autocorrelation, if necessary.
 
```{r set up df_model}
df_model <- df_summary %>% #create factors for year and domain
  mutate(YEAR = as.factor(YEAR), 
         DOMAIN = as.factor(DOMAIN))
 
# convert lat long to distances
# x <- mapproject(df_model$LON, df_model$LAT, "albers", param=c(55.9, 60.8))
# df_model$x <- x$x
# df_model$y <- x$y
```

```{r check spatial autocorr}
# Examine residuals for autocorrelation. First, we can do a quick graphical 
# exploration by plotting the residuals in space:
#spatial component follow along with lab 12 document 
spatial_df <- df_model
# Estimate means of CPUE by year:
mean_year <- gam(sum ~ YEAR, data=spatial_df, family = tw, method = "ML")
summary(mean_year) #p=1.858
mean_year$df.null
 
# We should examine autocorrelation for one year at a time because the spatial 
# autocorrelation, if present, should be evident within years only (because the 
# fish may have a quite different spatial distributions between years):
spatial_df$r <- resid(mean_year)   # Extract residuals
coordinates(spatial_df) <- c("LON","LAT")
# Note that this creates a more complex object that includes the data frame as
# a 'slot'. However, you can still use it like a data frame for most purposes.
j <- spatial_df$YEAR == "2014"  # Extract 2010 data only
bubble(spatial_df[j,], zcol="r")
# Note strong spatial patterns of all positive and all negative correlations!

# a 'slot'. However, you can still use it like a data frame for most purposes.
k <- spatial_df$YEAR == "2015"  # Extract 2010 data only
bubble(spatial_df[k,], zcol="r")

# a 'slot'. However, you can still use it like a data frame for most purposes.
l <- spatial_df$YEAR == "2016"  # Extract 2010 data only
bubble(spatial_df[l,], zcol="r")


j <- spatial_df$YEAR == "2014"   # Select year
r <- resid(mean_year)[j] 
d <- dist(coordinates(spatial_df)[j,])
d  # Large (88x88) matrix of pairwise distances (values below diagonal only!)
d <- as.vector(d)   # Need to convert to a vector for 'Variogram' function
# The vector d, contains only one set of pairwise distances (values from below
# the diagonal) and has length: n * (n-1) / 2, where n is the number of observations
# (see help files for 'dist' and 'Variogram')

# The 'Variogram' function from the 'nlme' package computes the semi-variogram,
#  given a variable (residuals in this case) and the corresponding pairwise 
#  distances 
SemiVar <- Variogram(r, d)
#plot(SemiVar, xlim=c(0,0.1)) # maybe dont plot so far this crashes the computer 

bins <- cut(SemiVar$dist, seq(0,0.1, by=0.005))  
plot(bins, SemiVar$variog) #looks like slight spatial autocorr, nothing too huge.
```
 
```{r CALANUS MODEL TESTING}
df_calanus <- df_model %>%
  filter(TAXON_NAME == "Calanus spp")%>%
  #filter(!TEMPERATURE < 0) %>%
  filter(!sum ==0)

m1 <- gam(sum ~  s(julian,k=4),
         data = df_calanus, method = "ML", family = Tweedie(p=1.84))

m2 <- gam(sum ~  s(TEMPERATURE, k=4),
         data = df_calanus, method = "ML", family = Tweedie(p=1.84))
 
m3 <- gam(sum ~    s(TEMPERATURE, k=4)+te(x, y, k=c(8,5)),
         data = df_calanus, method = "ML", family = Tweedie(p=1.84))

m4 <- gam(sum ~ YEAR + s(TEMPERATURE, k=4) + te(x, y,k=c(8,5), by = YEAR),
         data = df_calanus, method = "ML", family = Tweedie(p=1.84))

m5 <- gam(sum ~ YEAR + s(julian,k=4) + te(x, y,  by = YEAR,k=c(8,5)),
         data = df_calanus, method = "ML", family = Tweedie(p=1.84))

m6 <- gam(sum ~ YEAR +  s(TEMPERATURE, k=4) + s(julian,k=4) + te(x, y,  by = YEAR, k=c(8,5)),
          data = df_calanus, method = "ML", family = Tweedie(p=1.84))

AIC(m1,m2,m3,m4,m5,m6)
summary(m1)
summary(m2)
summary(m3)
summary(m4)
summary(m5)
summary(m6)

### ESTIMATE BEST MODEL USING REML
best_calanus <- gam(sum ~ YEAR +  s(TEMPERATURE, k=5) + s(julian, k =3) + te(LON, LAT,  by = YEAR),
          data = df_calanus, method = "REML", family = Tweedie(p=1.84))

```

```{r CALANUS BEST MODEL GOODNESS OF FIT}
summary(best_calanus)

#residuals x year
plot(df_calanus$YEAR, resid(best_calanus)); abline(h=0, lty=2)
plot(df_calanus$TEMPERATURE, resid(best_calanus)); abline(h=0, lty=2)
plot(df_calanus$julian, resid(best_calanus)); abline(h=0, lty=2)

b<-getViz(best_calanus) 
#### check spatial residuals 
ck1 <- check1D(b, "LON")
ck2 <- check1D(b, "LAT")
gridPrint(ck1, ck2, ncol = 2)
gridPrint(ck1 + l_dens(type = "cond", alpha = 0.8) + l_rug(alpha = 0.2), 
          ck2 + l_points() + l_rug(alpha = 0.2), layout_matrix = matrix(c(1, 1, 1, 2, 2), 1, 5)) #maybe over predicting in middle, high residuals. 
 
##### nice looking goodness of fits with mgcvViz
check(b,
      a.qq = list(method = "tnorm", 
                  a.cipoly = list(fill = "light blue")), 
      a.respoi = list(size = 0.5), 
      a.hist = list(bins = 10))

temp <- plot(sm(b, 1) )

temp + l_fitLine(colour = "red") + #l_rug(mapping = aes(x=x, y=y), alpha = 0.8) +
    l_ciLine(mul = 5, colour = "blue", linetype = 2) + 
    l_points(shape = 19, size = 1, alpha = 0.1) + 
    theme_classic()   

julian <- plot(sm(b, 2) )

julian + l_fitLine(colour = "red") + #l_rug(mapping = aes(x=x, y=y), alpha = 0.8) +
    l_ciLine(mul = 5, colour = "blue", linetype = 2) + 
    l_points(shape = 19, size = 1, alpha = 0.1) + 
    theme_classic()   

calanus_temp<-temp + l_fitLine(colour = "red") + #l_rug(mapping = aes(x=x, y=y), alpha = 0.8) +
    l_ciLine(mul = 5, colour = "blue", linetype = 2) + 
    l_points(shape = 19, size = 1, alpha = 0.1) + 
    theme_classic()  +
  ggtitle("Calanus")

#visualize spatial uncertainty
vis.gam(best_calanus, view = c("LON", "LAT"), 
        plot.type = "persp", se = 2)
#rotate
vis.gam(best_calanus, view = c("LON", "LAT"), 
        plot.type = "persp", se = 2, theta = 135)
#plot map
# Make plot with 5% extrapolation
vis.gam(best_calanus, view = c("LON", "LAT"), 
        plot.type = "contour", too.far = 0.05)

# Overlay data
points(df_model$LON,df_model$LAT)
```

```{r PSEUDOCALANUS MODEL TESTING}
df_psuedocalanus <- df_model %>%
  filter(TAXON_NAME == "Pseudocalanus spp.") %>%
  filter(!sum == 0)


ps1 <- gam(sum ~  s(julian,k=4),
         data = df_psuedocalanus, method = "ML", family = Tweedie(p=1.84))

ps2 <- gam(sum ~  s(TEMPERATURE, k=4),
         data = df_psuedocalanus, method = "ML", family = Tweedie(p=1.84))
 
ps3 <- gam(sum ~    s(TEMPERATURE, k=4)+te(x, y, k=c(8,5)),
         data = df_psuedocalanus, method = "ML", family = Tweedie(p=1.84))

ps4 <- gam(sum ~ YEAR + s(TEMPERATURE, k=4) + te(x, y,k=c(8,5), by = YEAR),
         data = df_psuedocalanus, method = "ML", family = Tweedie(p=1.84))

ps5 <- gam(sum ~ YEAR + s(julian,k=4) + te(x, y,  by = YEAR,k=c(8,5)),
         data = df_psuedocalanus, method = "ML", family = Tweedie(p=1.84))

ps6 <- gam(sum ~ YEAR +  s(TEMPERATURE, k=4) + s(julian,k=4) + te(x, y,  by = YEAR, k=c(8,5)),
          data = df_psuedocalanus, method = "ML", family = Tweedie(p=1.84))
  

AIC(ps1,ps2,ps3,ps4,ps5,ps6)
summary(ps1)
summary(ps2)
summary(ps3)
summary(ps4)
summary(ps5)
summary(ps6)
 
### ESTIMATE BEST MODEL USING REML
best_ps <- gam(sum ~ YEAR +  s(TEMPERATURE, k=4) + s(julian,k=4) + te(LON, LAT,  by = YEAR, k=c(8,5)),data = df_psuedocalanus, method = "REML", family = Tweedie(p=1.84))
  
```

```{r PSEUDOCALANUS BEST MODEL GOODNESS OF FIT}

#residuals x year
plot(df_psuedocalanus$YEAR, resid(ps5)); abline(h=0, lty=2)
plot(df_psuedocalanus$TEMPERATURE, resid(ps5)); abline(h=0, lty=2)
plot(df_psuedocalanus$julian, resid(ps5)); abline(h=0, lty=2)

p<-getViz(best_ps)
#### check residuals 
ck1 <- check1D(p, "LON")
ck2 <- check1D(p, "LAT")
gridPrint(ck1, ck2, ncol = 2)
gridPrint(ck1 + l_dens(type = "cond", alpha = 0.8) + l_rug(alpha = 0.2), 
          ck2 + l_points() + l_rug(alpha = 0.2), layout_matrix = matrix(c(1, 1, 1, 2, 2), 1, 5)) #maybe over predicting in middle, high residuals. 

##### nice looking goodness of fits with mgcvViz
check(p,
      a.qq = list(method = "tnorm", 
                  a.cipoly = list(fill = "light blue")), 
      a.respoi = list(size = 0.5), 
      a.hist = list(bins = 10))

temp <- plot( sm(p, 1) )

temp + l_fitLine(colour = "red") + #l_rug(mapping = aes(x=x, y=y), alpha = 0.8) +
    l_ciLine(mul = 5, colour = "blue", linetype = 2) + 
    l_points(shape = 19, size = 1, alpha = 0.1) + theme_classic()  

pseudo_temp <- temp + l_fitLine(colour = "red") + #l_rug(mapping = aes(x=x, y=y), alpha = 0.8) +
    l_ciLine(mul = 5, colour = "blue", linetype = 2) + 
    l_points(shape = 19, size = 1, alpha = 0.1) + theme_classic()  +
  ggtitle("Pseudocalanus")

julian_ps <- plot(sm(p, 2) )

julian_ps + l_fitLine(colour = "red") + #l_rug(mapping = aes(x=x, y=y), alpha = 0.8) +
    l_ciLine(mul = 5, colour = "blue", linetype = 2) + 
    l_points(shape = 19, size = 1, alpha = 0.1) + theme_classic()  

pseudo_julian <- julian_ps + l_fitLine(colour = "red") + #l_rug(mapping = aes(x=x, y=y), alpha = 0.8) +
    l_ciLine(mul = 5, colour = "blue", linetype = 2) + 
    l_points(shape = 19, size = 1, alpha = 0.1) + theme_classic()  +
  ggtitle("Pseudocalanus") 
 
#visualize spatial uncertainty
vis.gam(ps5, view = c("LON", "LAT"), 
        plot.type = "persp", se = 2)
#rotate
vis.gam(ps5, view = c("LON", "LAT"), 
        plot.type = "persp", se = 2, theta = 135)
#plot map
# Make plot with 5% extrapolation
vis.gam(m5, view = c("LON", "LAT"), 
        plot.type = "contour", too.far = 0.05)

# Overlay data
points(df_model$LON,df_model$LAT)
```

```{r save data plots to pdf for talk}
pdf("final_project_figures/temperature_map.pdf")
temperature
dev.off()
 
# #save temp plots
# pdf("final_project_figures/calanus_point_map.pdf")
# cal_point_map
# dev.off()
# 
# pdf("final_project_figures/pseudocalanus_point_map.pdf")
# ps_point_map
# dev.off()
# 
#  
point_both_sp<-ggarrange(cal_point_map,ps_point_map, labels = c("A", "B"))

pdf("final_project_figures/both_sp_point_map.pdf")
point_both_sp
dev.off()

#save temp plots
pdf("final_project_figures/calanus_gam_temperature.pdf")
calanus_temp
dev.off()

pdf("final_project_figures/pseudocalanus_gam_temperature.pdf")
pseudo_temp
dev.off()
```
 
 
 
```{r early and late temperature model and plot for smooth temperature}
df_temperature_model<-df_summary %>% filter(!TEMPERATURE <0.01) %>%
  mutate(YEAR = as.factor(YEAR)) %>% 
  filter(case_when(YEAR == 2018 ~ LAT < 63,
                   TRUE ~TRUE ))

temp <- gam(TEMPERATURE ~ YEAR+ te(LON, LAT,  by = YEAR, k=c(8,5)) + s(julian, k =3) ,  
         data = df_temperature_model, method = "REML", family = Gamma(link = "log") )
 
#first need to generate smooth temperature to predict Calanus on. do that based on temp model. 
new_temp_early <- expand_grid(YEAR = c("2014", "2015", "2016", "2018"),
                        julian = 150, 
                        LAT = seq(min(df_temperature_model$LAT), max(df_temperature_model$LAT), length=100),
                        LON = seq(min(df_temperature_model$LON), max(df_temperature_model$LON), length =100 ))
new_temp_early$temp<-predict(temp,new_temp_early) 


new_temp_late <- expand_grid(YEAR = c("2014", "2015", "2016", "2018"),
                        julian = 250, 
                        LAT = seq(min(df_temperature_model$LAT), max(df_temperature_model$LAT), length=100),
                       LON = seq(min(df_temperature_model$LON), max(df_temperature_model$LON), length =100 ))
new_temp_late$temp<-predict(temp,new_temp_late) 

#filter temperature script appropriately to remove extrapolation
#have to do this by year... 
temperature_filter_by_year<-function(year,temp_df){
points <- df_temperature_model %>% 
  filter(YEAR == year) %>%
  st_as_sf(coords = c("LON","LAT"))#, crs=4326)

polygon <- st_union(points) %>%  # unite the points to 1 object
  st_convex_hull() %>% # calculate the convex hull
  st_as_sf()

# verify the results by drawing a map
# ggplot() +
#   geom_sf(data = points, col = "red", size = 2, pch = 4) +
#   geom_sf(data = polygon, col = "grey45", fill = NA)

temperature_sf<-st_as_sf(temp_df %>% filter(YEAR == year),
                         coords = c("LON","LAT"))

temperature_df <-st_intersection(temperature_sf, polygon) %>%
  data.frame() %>%
  separate(geometry, into =c( "LON","LAT" ), sep = "," ) %>%
  mutate(LON = as.numeric(str_sub(LON, 3)),
         LAT = as.numeric(str_sub(LAT, 1,-2)))
return(temperature_df)
}

#run function for each year and bind together
temp_2014<-temperature_filter_by_year(year = "2014", temp_df = new_temp_early)
temp_2015<-temperature_filter_by_year(year = "2015", temp_df = new_temp_early)
temp_2016<-temperature_filter_by_year(year = "2016", temp_df = new_temp_early)
temp_2018<-temperature_filter_by_year(year = "2018", temp_df = new_temp_early)

temperature_df_filtered_early<-rbind(temp_2014,temp_2015,temp_2016,temp_2018) %>%
  filter(!temp >log(12))

temp_2014<-temperature_filter_by_year(year = "2014", temp_df = new_temp_late)
temp_2015<-temperature_filter_by_year(year = "2015", temp_df = new_temp_late)
temp_2016<-temperature_filter_by_year(year = "2016", temp_df = new_temp_late)
temp_2018<-temperature_filter_by_year(year = "2018", temp_df = new_temp_late)

temperature_df_filtered_late<-rbind(temp_2014,temp_2015,temp_2016,temp_2018) %>%
  filter(!temp >log(12))
 
#https://stackoverflow.com/questions/57474365/how-to-draw-polygon-for-lots-of-lat-long-coordinates-and-calculate-surface-are

ggplot(ak) +
   geom_contour_filled(data= temperature_df_filtered_early,
        aes(LON, LAT, z= exp(temp))) +
  geom_sf( fill= "white") +
  #geom_point(data = df_temperature_model, aes(x=LON, y=LAT)) +
  facet_wrap(~YEAR ) +
      scale_fill_viridis_d(name="temp") +
      theme( strip.text = element_text(size =12),
        legend.key.size = unit(0.5, 'cm'),
        legend.position = "bottom",
        legend.text=element_text(size=7),
        legend.title = element_text(size=7),
        plot.caption = element_text(hjust = 0, size =10),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

```{r Plot best models on a map CALANUS, early}
#### take df from temperature smooth and predict Calanus abundance. 
cal_predict <- temperature_df_filtered_early %>% 
  mutate(julian = 150) %>%
  dplyr::rename(TEMPERATURE = temp)

pred <- predict(best_calanus, cal_predict, se.fit=TRUE)

cal_predict <- cbind(cal_predict, data.frame(pred[1]), data.frame(pred[2])) 

#filter out areas where it is clearly extrapolating. 
 cal<-cal_predict %>%
   filter(!se.fit>1.5) # DO A BETTER JOB AT FILTERING THESE FOR WHERE THERE IS LITTLE DATA, IE 2016 AND NORTHERN PORTION OF 2017
 
calanus_pred<-ggplot(ak) +
   geom_contour_filled(data= cal_predict,
        aes(LON, LAT, z= fit)) + 
  geom_sf( fill= "white") +
# geom_sf(data = ortiz_shp, fill = NA) + 
  #geom_point(data = df_model, aes(x=LON, y=LAT)) + 
  facet_wrap(~YEAR) + 
      scale_fill_viridis_d(name="Calanus log(CPUE)") +
      theme( strip.text = element_text(size =12),
        legend.key.size = unit(0.5, 'cm'),
        legend.position = "bottom", 
        legend.text=element_text(size=7),
        legend.title = element_text(size=7),
        plot.caption = element_text(hjust = 0, size =10),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())  

calanus_pred
```

```{r Plot best models on a map CALANUS, late}
#### take df from temperature smooth and predict Calanus abundance. 
cal_predict <- temperature_df_filtered_late %>% 
  mutate(julian = 250) %>%
  dplyr::rename(TEMPERATURE = temp)

pred <- predict(best_calanus, cal_predict, se.fit=TRUE)

cal_predict <- cbind(cal_predict, data.frame(pred[1]), data.frame(pred[2])) 

#filter out areas where it is clearly extrapolating. 
 cal<-cal_predict %>%
   filter(!se.fit>1.5) # DO A BETTER JOB AT FILTERING THESE FOR WHERE THERE IS LITTLE DATA, IE 2016 AND NORTHERN PORTION OF 2017
 
calanus_pred_late<-ggplot( ak) +
   geom_contour_filled(data= cal,
        aes(LON, LAT, z= fit)) + 
  geom_sf( fill= "white") +
 #geom_point(data = df_model, aes(x=LON, y=LAT)) + 
  facet_wrap(~YEAR) + 
      scale_fill_viridis_d(name="Calanus log(CPUE), Julian day = 250  ") +
      theme( strip.text = element_text(size =12),
        legend.key.size = unit(0.5, 'cm'),
        legend.position = "bottom", 
        legend.text=element_text(size=7),
        legend.title = element_text(size=7),
        plot.caption = element_text(hjust = 0, size =10),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())  

calanus_pred_late
```

```{r Plot best models on a map PSUEDOCALANUS, early}
PScal_predict <- temperature_df_filtered_early %>% 
    mutate(julian = 150) %>%
  dplyr::rename(TEMPERATURE = temp)

PScal_pred <- predict(best_ps, PScal_predict, se.fit=TRUE)

PScal_predict <- cbind(PScal_predict, data.frame(PScal_pred[1]), data.frame(PScal_pred[2])) 

#filter out areas where it is clearly extrapolating. 
PScal<-PScal_predict %>%
  filter(!se.fit>1.5)#!est>8.5)

ps_map_early<-ggplot(ak) +
   geom_contour_filled(data= PScal,
        aes(LON, LAT, z= fit)) + 
  geom_sf( fill= "white") + 
  # geom_sf(data = ortiz_shp, fill= NA) + 
  facet_wrap(~YEAR) + 
      scale_fill_viridis_d(name="Pseudocalanus log(CPUE), Early") +
      theme( strip.text = element_text(size =12),
        legend.key.size = unit(0.5, 'cm'),
        legend.position = "bottom", 
        legend.text=element_text(size=7),
        legend.title = element_text(size=7),
        plot.caption = element_text(hjust = 0, size =10),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())  

ps_map_early

```

```{r Plot best models on a map PSUEDOCALANUS, late}
PScal_predict <- temperature_df_filtered_late %>% 
    mutate(julian = 250) %>%
  dplyr::rename(TEMPERATURE = temp)

PScal_pred <- predict(best_ps, PScal_predict, se.fit=TRUE)

PScal_predict <- cbind(PScal_predict, data.frame(PScal_pred[1]), data.frame(PScal_pred[2])) 

#filter out areas where it is clearly extrapolating. 
PScal<-PScal_predict %>%
  filter(!se.fit>1.5)#!est>8.5)

ps_map_late<-ggplot(ak) +
   geom_contour_filled(data= PScal,
        aes(LON, LAT, z= fit)) + 
  geom_sf( fill= "white") + 
  # geom_sf(data = ortiz_shp, fill= NA) + 
  facet_wrap(~YEAR) + 
      scale_fill_viridis_d(name="Pseudocalanus log(CPUE)") +
      theme( strip.text = element_text(size =12),
        legend.key.size = unit(0.5, 'cm'),
        legend.position = "bottom", 
        legend.text=element_text(size=7),
        legend.title = element_text(size=7),
        plot.caption = element_text(hjust = 0, size =10),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())  

ps_map_late

```

```{r PLOT COEFFICIENTS}

cal_summary<-summary(best_calanus)
cal_output<-data.frame(coeff=cal_summary$p.coeff)
cal_output$se<- cal_summary$se[1:4]
cal_output$species <- "Calanus"

ps_summary<- summary(best_ps)
ps_output<-data.frame(coeff=ps_summary$p.coeff)
ps_output$se<- ps_summary$se[1:4]
ps_output$species <- "Pseudocalanus"

coeffdf<-rbind(ps_output, cal_output) 

 year_estimates_df <- coeffdf %>%
   rownames_to_column() %>% 
  dplyr::select(-se) %>%
  spread(rowname, coeff) %>%
  transmute(Calanus.year2014 = `(Intercept)`,
            Calanus.year2015 = `(Intercept)`+YEAR2015,
            Calanus.year2016 = `(Intercept)`+YEAR2016,
            Calanus.year2018 = `(Intercept)`+YEAR2018,
            
            Pseudo.year2014 = `(Intercept)1`,
            Pseudo.year2015 = `(Intercept)1`+YEAR20151,
            Pseudo.year2016 = `(Intercept)1`+YEAR20161,
            Pseudo.year2018 = `(Intercept)1`+YEAR20181) %>%
   gather(1:ncol(.), key = "year", value = "log_estimate") %>%
   filter(!is.na(log_estimate)) %>%
   cbind(coeffdf %>% dplyr::select(-coeff)) %>%
   separate(year, into = c("delete", "year"))
library(PNWColors)
 
year_CPUE<-ggplot(year_estimates_df, aes(x=year, y = log_estimate, group = species, fill = year)) +
  geom_bar(stat = 'identity', color = "black"  ) +
  geom_errorbar(aes(ymin = log_estimate - se, ymax = log_estimate + se), width = 0.2) +
  facet_wrap(~species) + 
  scale_fill_manual(values = pnw_palette("Bay", n=4, "continuous")) +
  guides(fill = "none")+
  theme(axis.text.x =  element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab( " ") +
   ylab("Log Estimate Mean CPUE")
year_CPUE

pdf("final_project_figures/year_cpue.pdf")
year_CPUE
dev.off()
```


```{r save pretty maps}
pdf("final_project_figures/PS_map.pdf")
ggarrange(ps_map_early,ps_map_late, labels = c("A", "B"))
dev.off()

pdf("final_project_figures/CAL_map.pdf")
ggarrange(calanus_pred,calanus_pred_late, labels = c("A", "B"))
dev.off()

# pdf("final_project_figures/PS_map_late.pdf")
# ps_map_late
# dev.off()
# 
# #save plot
# pdf("final_project_figures/Calanus_map_early.pdf")
# calanus_pred_early
# dev.off()
# 
# pdf("final_project_figures/Calanus_map_late.pdf")
# calanus_pred_late
# dev.off()
```

## Acknowledgements
Thank you to Courtney Weiss and Molly Payne for providing comments. Thank you to Dave Kimmel for providing data and for reviewing ideas with me during the project. 

## References
- Baumgartner, M. F., Lysiak, N. S., Esch, H. C., Zerbini, A. N., Berchok, C. L., & Clapham, P. J. (2013). Associations between North Pacific right whales and their zooplanktonic prey in the southeastern Bering Sea. Marine Ecology Progress Series, 490, 267-284.
- Cheung, W.W.L., T.L. Frölicher (2020) Marine heatwaves exacerbate climate change impacts for fisheries in the northeast Pacific. Sci Rep 10, 6678 https://doi.org/10.1038/s41598-020-63650-z
- Danielson, S., Hedstrom, K., Aagaard, K., Weingartner, T., & Curchitser, E. (2012). Wind-induced reorganization of the Bering shelf circulation. Geophysical Research Letters, 39(8), L08601. https://doi.org/10.1029/2012gl051231
- Fisher, M.C., S.K. Moore, S.L. Jardine, J.R. Watson, & J.F. Samhouri (2021) Climate shock effects and mediation in fisheries. Proceedings of the National Academy of Sciences 118, no. 2.
- Frölicher, T.L., E.M. Fischer, & N Gruber. (2020) Marine heatwaves under global warming." Nature 560.7718 (2018): 360-364.
- Hansen, B. B., Grøtan, V., Herfindal, I., & Lee, A. M. (2020) The Moran effect revisited: spatial population synchrony under global warming. Ecography, 43(11), 1591-1602.
- Siddon, E.C., T. Kristiansen, F.J. Mueter, K.K. Holsman, R.A. Heintz, and E.V. Farley. (2013) Spatial match-mismatch between juvenile fish and prey provides a mechanism for recruitment variability across contrasting climate conditions in the eastern Bering Sea. PLoS One 8, no. 12 (2013): e84526.
- Springer, A. M., Roseneau, D. G., Lloyd, D. S., McRoy, C. P., & Murphy, E. C. (1986). Seabird responses to fluctuating prey availability in the eastern Bering Sea. Marine Ecology Progress Series, 1-12.